--
title: "PGA Tour Data Overview"
output: html_document
---

This report examines the structure, coverage, and completeness of the 2025 PGA Tour hole-level data used in our next-hole performance analysis.

## Data coverage

```{r}
results <- readRDS("holes_2025_raw.rds")

failed <- names(results)[
  sapply(results, function(x) is.null(x) || nrow(x) == 0)
]

data.frame(
  total_events = length(results),
  events_with_data = length(results) - length(failed)
)

show(failed)
```

## Structure of the hole data

Each row represents one player's score on one hole in one round. The Sentry is used below as an example.

```{r}
holes <- results[["The Sentry"]]

data.frame(rows = nrow(holes), columns = ncol(holes))
head(holes)
```

## Round completeness across all events

```{r}
library(dplyr)
library(DBI)
library(RSQLite)

all_holes <- bind_rows(results, .id = "event_query")

con <- dbConnect(SQLite(), ":memory:")
dbWriteTable(con, "all_holes", all_holes)

dbGetQuery(con, "
  SELECT n_holes, COUNT(*) AS n_player_rounds
  FROM (
    SELECT event_query, player_id, round, COUNT(*) AS n_holes
    FROM all_holes
    GROUP BY event_query, player_id, round
  )
  GROUP BY n_holes
  ORDER BY n_holes
")

dbDisconnect(con)
```

This check counts recorded rows per player-round. An incomplete round
may reflect a withdrawal or missing data. Eighteen rows alone do not
guarantee eighteen distinct holes.