---
title: "PGA Tour Momentum Project"
format: html
editor: visual
---


## Inspect the raw data

```{r}
results <- readRDS("holes_2025_raw.rds")

# Number of events and events with no data
length(results)

failed <- names(results)[
  sapply(results, function(x) is.null(x) || nrow(x) == 0)
]
failed
```

### Example: The Sentry

Each row records one player's score on one hole in one round.

```{r}
holes <- results[["The Sentry"]]

dim(holes)
names(holes)
head(holes)
str(holes)
```

### Check incomplete rounds

```{r}
library(DBI)
library(RSQLite)

# Use a temporary database to inspect the raw event
inspection_con <- dbConnect(SQLite(), ":memory:")
dbWriteTable(inspection_con, "holes", holes)

dbGetQuery(inspection_con, "
  SELECT player_id, player_name, round, COUNT(*) AS n_holes
  FROM holes
  GROUP BY player_id, player_name, round
  HAVING COUNT(*) != 18
")
```

Davis Riley's fourth round contains nine recorded holes, numbered 10–18.
This may reflect a withdrawal or missing data; the hole records alone
do not establish the cause.

```{r}
dbGetQuery(inspection_con, "
  SELECT player_name, hole, par, score
  FROM holes
  WHERE player_id = '10058' AND round = 4
  ORDER BY hole
")

dbDisconnect(inspection_con)
```

## Build the database

R reads the saved golfastr responses and imports them into SQLite. Filtering, course proxies, previous-hole features, and difficulty controls are computed in `sql/analysis.sql`. Python can later query the same `analysis` view for modeling.

```{r}
source("02_build_database.R")
con <- dbConnect(RSQLite::SQLite(), "pga_2025.sqlite")


results <- readRDS("holes_2025_raw.rds")

```

## Check round sizes and extra rounds

```{r}
dbGetQuery(con, "
  SELECT n_holes, COUNT(*) AS n_player_rounds
  FROM (
    SELECT event_query, round, player_id, COUNT(*) AS n_holes
    FROM clean GROUP BY event_query, round, player_id
  ) GROUP BY n_holes ORDER BY n_player_rounds DESC LIMIT 20
")
dbGetQuery(con, "
  SELECT event_query, round, COUNT(*) AS n_rows
  FROM clean WHERE round = 5 GROUP BY event_query, round
")
```

## Check course proxies

The course key concatenates pars in hole-number order. Different courses can
share the same par sequence, so this is only a proxy. Eligibility preserves the
original 18-row rule; this additional check flags duplicate or missing holes.

```{r}
dbGetQuery(con, "
  SELECT event_query, round, COUNT(DISTINCT course_key) AS n_courses
  FROM round_courses GROUP BY event_query, round
  HAVING COUNT(DISTINCT course_key) > 1 ORDER BY event_query, round
")
dbGetQuery(con, "
  SELECT event_query, round, player_id
  FROM complete_rounds GROUP BY event_query, round, player_id
  HAVING COUNT(DISTINCT hole) != 18 OR MIN(hole) != 1 OR MAX(hole) != 18
")
```

## Analysis features

`prev_birdie` means birdie **or better**, matching the original `<= -1` rule.
The lag resets at each nine, so holes 1 and 10 have no previous-hole value.
This uses hole-number order, not verified playing order for shotgun or split-tee
starts. `hole_diff` excludes the current player's score; `player_skill` excludes
the current event. Undefined controls are SQL NULL via `NULLIF`.

```{r}
dbGetQuery(con, "
  SELECT COUNT(*) AS n_holes,
    SUM(prev_to_par IS NULL) AS missing_previous,
    SUM(hole_diff IS NULL) AS missing_hole_diff,
    SUM(player_skill IS NULL) AS missing_player_skill,
    MIN(hole_diff) AS min_hole_diff, MAX(hole_diff) AS max_hole_diff,
    MIN(player_skill) AS min_player_skill, MAX(player_skill) AS max_player_skill
  FROM analysis
")
dbGetQuery(con, "
  SELECT prev_birdie, COUNT(*) AS n, AVG(to_par) AS mean_next
  FROM analysis WHERE prev_birdie IS NOT NULL
  GROUP BY prev_birdie ORDER BY prev_birdie
")
```

## Validate score labels

```{r}
dbGetQuery(con, "
  SELECT score_type, COUNT(*) AS n FROM analysis GROUP BY score_type
")
dbGetQuery(con, "
  SELECT score, par, score_type, COUNT(*) AS n
  FROM analysis WHERE to_par <= -3 GROUP BY score, par, score_type
")
dbGetQuery(con, "
  WITH checked AS (
    SELECT *, CASE
      WHEN score_type = 'OTHER' THEN to_par < 4
      ELSE to_par != CASE score_type
        WHEN 'DOUBLE_EAGLE' THEN -3 WHEN 'EAGLE' THEN -2
        WHEN 'BIRDIE' THEN -1 WHEN 'PAR' THEN 0 WHEN 'BOGEY' THEN 1
        WHEN 'DOUBLE_BOGEY' THEN 2 WHEN 'TRIPLE_BOGEY' THEN 3
      END
    END AS mismatch FROM analysis
  )
  SELECT event_query, round, SUM(mismatch) AS bad, COUNT(*) AS total
  FROM checked GROUP BY event_query, round
  HAVING SUM(mismatch) > 0 ORDER BY bad DESC LIMIT 30
")
dbDisconnect(con)
```
