---
title: "PGA Tour Momentum Project"
format:
  html:
    code-fold: true
editor: source
---

## The Purpose of the Project

As a ncaa d1 college golfer, I was interested in whether the momentum / hot hand feeling during a round actually shows up in the data. A birdie feels like it should carry over. But maybe the next hole is just easier, or the player is simply having a good day. Our Research Question: **Is making a birdie or better associated with a better score on the next hole, after accounting for player ability and hole difficulty?** 

## Data and a little golf background

The 2025 PGA Tour hole-level data was retrieved through the R package `golfastr`, using `load_leaderboard()` to identify events and `load_holes()` to retrieve their hole records. The saved dataframes are in `holes_2025_raw.rds`, so the analysis does not need to download them again.

Each row represents one player's score on one hole in one tournament round. A round normally has 18 holes. Par is the reference number of strokes for a hole. The labels below describe a player's score relative to par. Lower scores are better.

| Score label | Strokes relative to par | Example on a par-5 hole |
|---|---:|---:|
| Double eagle (albatross) | −3 | 2 |
| Eagle | −2 | 3 |
| Birdie | −1 | 4 |
| Par | 0 | 5 |
| Bogey | +1 | 6 |
| Double bogey | +2 | 7 |
| Triple bogey | +3 | 8 |
| Other | A separate category checked against the recorded score | — |

For the analysis, we calculate `to_par = score - par` directly from the numeric scores.

R loads the data into SQLite, SQL prepares the analysis variables, and Python modeling is the next step. The sections below check the data and prepare those variables; they do not yet provide an adjusted estimate of momentum.


## Build the database

R reads the saved golfastr responses and imports them into SQLite. Filtering, course proxies, previous-hole features, and difficulty controls are computed in `sql/analysis.sql`. Python can later query the same `analysis` view for modeling.

```{r}
library(DBI)
library(RSQLite)

pga_db <- dbConnect(SQLite(), "pga_2025.sqlite")

```

## Check round sizes and extra rounds

A player can withdraw before finishing a round, and data can also be missing. The first table counts how many player-rounds have each number of recorded holes across the included events. The second checks records labeled as round 5. Our current sample keeps rounds 1–4 with 18 recorded rows; other complete rounds from the same player remain eligible.

```{r}
dbGetQuery(pga_db, "
  SELECT n_holes, COUNT(*) AS n_player_rounds
  FROM (
    SELECT event_query, round, player_id, COUNT(*) AS n_holes
    FROM clean GROUP BY event_query, round, player_id
  ) GROUP BY n_holes ORDER BY n_player_rounds DESC LIMIT 20
")
dbGetQuery(pga_db, "
  SELECT event_query, round, COUNT(*) AS n_rows
  FROM clean WHERE round = 5 GROUP BY event_query, round
")
```

## Check course proxies

Some events use more than one course. We use the sequence of pars from holes 1–18 as a rough course identifier, so players on different layouts are not automatically compared as if they played the same hole. Different courses can share this sequence, so it is an imperfect proxy.

The first query finds event-rounds with multiple par sequences. The second checks whether an 18-row player-round actually contains holes 1–18 without duplicates. This check reports problems; it does not remove them automatically.

```{r}
dbGetQuery(pga_db, "
  SELECT event_query, round, COUNT(DISTINCT course_key) AS n_courses
  FROM round_courses GROUP BY event_query, round
  HAVING COUNT(DISTINCT course_key) > 1 ORDER BY event_query, round
")
dbGetQuery(pga_db, "
  SELECT event_query, round, player_id
  FROM complete_rounds GROUP BY event_query, round, player_id
  HAVING COUNT(DISTINCT hole) != 18 OR MIN(hole) != 1 OR MAX(hole) != 18
")
```

## Analysis features

We calculate each score relative to par: `to_par = score - par`. A birdie is −1, par is 0, and a bogey is +1.

- `prev_birdie`: whether the preceding hole was a birdie or better, including eagles.
- `hole_diff`: the other players' average score relative to par on the same event-round, course proxy, and hole. Lower values suggest an easier hole.
- `player_skill`: the player's average score relative to par in the other included events. Lower values suggest better scoring, though this is a rough measure rather than a full ability estimate.

For now, consecutive-hole comparisons stay within each nine, so holes 1 and 10 have no previous-hole value. Hole-number order is not always actual playing order; this rule avoids connecting the two nines but also omits valid transitions between them. A control is left missing when there are no other observations to calculate it from.

The first query checks sample size and missing controls. The second is a simple, unadjusted comparison of next-hole scores. Differences here could still reflect the players or holes involved.

```{r}
dbGetQuery(pga_db, "
  SELECT COUNT(*) AS n_holes,
    SUM(prev_to_par IS NULL) AS missing_previous,
    SUM(hole_diff IS NULL) AS missing_hole_diff,
    SUM(player_skill IS NULL) AS missing_player_skill,
    MIN(hole_diff) AS min_hole_diff, MAX(hole_diff) AS max_hole_diff,
    MIN(player_skill) AS min_player_skill, MAX(player_skill) AS max_player_skill
  FROM analysis
")
dbGetQuery(pga_db, "
  SELECT prev_birdie, COUNT(*) AS n, AVG(to_par) AS mean_next
  FROM analysis WHERE prev_birdie IS NOT NULL
  GROUP BY prev_birdie ORDER BY prev_birdie
")
```

## Validate score labels

Finally, we compare the recorded labels with `score - par`. For example, a row labeled BIRDIE should have a value of −1. The last query lists event-rounds with label mismatches for further inspection.

```{r}
dbGetQuery(pga_db, "
  SELECT score_type, COUNT(*) AS n FROM analysis GROUP BY score_type
")
dbGetQuery(pga_db, "
  SELECT score, par, score_type, COUNT(*) AS n
  FROM analysis WHERE to_par <= -3 GROUP BY score, par, score_type
")
dbGetQuery(pga_db, "
  WITH checked AS (
    SELECT *, CASE
      WHEN score_type = 'OTHER' THEN to_par < 4
      ELSE to_par != CASE score_type
        WHEN 'DOUBLE_EAGLE' THEN -3 WHEN 'EAGLE' THEN -2
        WHEN 'BIRDIE' THEN -1 WHEN 'PAR' THEN 0 WHEN 'BOGEY' THEN 1
        WHEN 'DOUBLE_BOGEY' THEN 2 WHEN 'TRIPLE_BOGEY' THEN 3
      END
    END AS mismatch FROM analysis
  )
  SELECT event_query, round, SUM(mismatch) AS bad, COUNT(*) AS total
  FROM checked GROUP BY event_query, round
  HAVING SUM(mismatch) > 0 ORDER BY bad DESC LIMIT 30
")
dbDisconnect(pga_db)
```